---
title: "Custom Tests"
author: "Shawn Garbett"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tangram)
```

## The Problem

Suppose one wanted custom tests for a demographics table. Instead of the provided
non-parametric tests one wanted a *t*-test?


```{r defaults}
tangram(sex ~ bili + drug, style="nejm", test=TRUE, data=pbc)
```

One can pass a function into the test. However, this function has to match 
the type dispatching that is done by the normal test functions.

```{r new_test}
mytest <- function(row, col, cell_style, ...)
{
  datar <- row$data
  datac <- col$data
  rtype <- hmisc_data_type(datar)
  ctype <- hmisc_data_type(datac)
  
  if(rtype == "Numerical" && ctype == "Numerical")
  {
     stat <- suppressWarnings(
       cor.test(datar, datac,
         alternate="two.sided", method="spearman",
         na.action=na.omit, exact=FALSE)
      )
     hmisc_spearman(
       stat$statistic,
       render_f(estimate, row$format),
       hmisc_p(stat$p.value, NULL)
     )

  } else if(rtype =="Categorical" && ctype == "Categorical")
  {
    grid <- table(as.categorical(datar), as.categorical(datac),useNA="no")
    if(is.na(colnames(grid)[ncol(grid)])) grid <- grid[,1:(ncol(grid)-1)]

    validcol  <- which(!apply(grid,2,FUN = function(x){all(x == 0)}))
    validrow  <- which(!apply(grid,1,FUN = function(x){all(x == 0)}))
    stat      <- if(length(validrow) < 2 || length(validcol) < 2) NA else                                     suppressWarnings(chisq.test(grid[validrow,validcol], correct=FALSE))
    if(any(is.na(stat))) cell("NA") else
      hmisc_chi2(
        render_f(stat$statistic, 2),
        stat$parameter,
        hmisc_p(stat$p.value, NULL)
      )
  } else { # Num X Cat or Cat x Num
    stat <- if(rtype == "Numerical")
    {
      suppressWarnings(tangram:::spearman2.default(datac, datar, na.action=na.retain))
    } else {
      suppressWarnings(tangram:::spearman2.default(datar, datac, na.action=na.retain))
    }
    hmisc_fstat(f   = render_f(stat['F'], "%.2f"),
                df1 = stat['df1'],
                df2 = stat['df2'],
                p   = hmisc_p(stat['P'], NULL))
    }
} 


tangram(sex ~ bili + drug, style="nejm", test=mytest, data=pbc)
```

That's a lot of code. This is what the logic of the default test function in the hmisc
transform. It's making a lot of decisions, and now we wish to change that. 

What we're going to do now, is if the number of categories is 2 apply a *t*-test instead.

```{r ttest}
mytest <- function(row, col, cell_style, ...)
{
  datar <- row$data
  datac <- col$data
  rtype <- hmisc_data_type(datar)
  ctype <- hmisc_data_type(datac)

  if(rtype == "Numerical" && ctype == "Numerical")
  {
     stat <- suppressWarnings(
       cor.test(datar, datac,
         alternate="two.sided", method="spearman",
         na.action=na.omit, exact=FALSE)
      )
     hmisc_spearman(
       stat$statistic,
       render_f(estimate, row$format),
       hmisc_p(stat$p.value, NULL)
     )

  } else if(rtype =="Categorical" && ctype == "Categorical")
  {
    grid <- table(as.categorical(datar), as.categorical(datac),useNA="no")
    if(is.na(colnames(grid)[ncol(grid)])) grid <- grid[,1:(ncol(grid)-1)]

    validcol  <- which(!apply(grid,2,FUN = function(x){all(x == 0)}))
    validrow  <- which(!apply(grid,1,FUN = function(x){all(x == 0)}))
    stat      <- if(length(validrow) < 2 || length(validcol) < 2) NA else
      suppressWarnings(chisq.test(grid[validrow,validcol], correct=FALSE))
    if(any(is.na(stat))) cell("NA") else
      hmisc_chi2(
        render_f(stat$statistic, 2),
        stat$parameter,
        hmisc_p(stat$p.value, NULL)
      )
  } else { # Num X Cat or Cat x Num
    if(rtype == "Categorical")
    {
      # Make it Num x Cat
      tmp   <- datac
      datac <- datar
      datar <- tmp
    }
    
    if(length(levels(datac)) == 2)
    {
      x <- datar[datac == levels(datac)[1]]
      y <- datar[datac == levels(datac)[2]]
      stat <- t.test(x, y)
      cell(paste0("*t* = ",
                  round(stat$statistic, 2),
                  ", ",
                  hmisc_p(stat$p.value, NULL),
                  "^4^"),
           class=c("statistics"),
           ...)
    } else {
      stat <- suppressWarnings(tangram:::spearman2.default(datar, datac, na.action=na.retain))
      hmisc_fstat(f   = render_f(stat['F'], "%.2f"),
                  df1 = stat['df1'],
                  df2 = stat['df2'],
                  p   = hmisc_p(stat['P'], NULL))
    }
  }
} 


tangram(sex ~ bili + drug, style="nejm", test=mytest, data=pbc,
        footnote="N is the number of non-missing value. ^1^Kruskal-Wallis. ^2^Pearson. ^3^Wilcoxon. ^4^*t*-test")
```
# Show Both

```{r both}
x <- tangram(sex ~ bili + drug, style="nejm", test=TRUE, data=pbc, footnote="N is the number of non-missing value. ^1^Kruskal-Wallis. ^2^Pearson. ^3^Wilcoxon. ^4^*t*-test")
y <- tangram(sex ~ bili + drug, style="nejm", test=mytest, data=pbc,
        ) %>% del_col(1:4)
cbind(x,y)
```
