---
title: "Matched Cohort Tables"
author: "Shawn Garbett, Kyle Rove"
date: '`r Sys.Date()`'
output:
  html_document: null
  pdf_document: null
header-includes:
- \usepackage[LGR,T1]{fontenc}
- \usepackage[utf8]{inputenc}
- \usepackage{textgreek}
- \usepackage{float}
- \usepackage[x11names,dvipsnames,table]{xcolor}
- \usepackage{boldline}
- \usepackage{multirow}
- \usepackage{colortbl}
- \usepackage{hhline}
- \usepackage{longtable}
- \usepackage{relsize}
email: shawn.garbett@vumc.org
vignette: |
  %\VignetteIndexEntry{Matched Cohort Examples}
  %\VignetteEngine{knitr::rmarkdown}
---

<style type="text/css">

body, td {
   font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 10px
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tangram)

library(tidyverse)
library(MatchIt)
library(geepack)
library(broom)
library(DescTools)
library(exact2x2)

pbc <- tangram::pbc

# This is necessary for changing size of chunks in LaTeX. Why isn't this patched in knitr?
# https://stackoverflow.com/questions/25646333/code-chunk-font-size-in-rmarkdown-with-knitr-and-latex/46526740
# ?highlight
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

pagebreak <- function() {
  if(knitr::is_latex_output())
    return("\\newpage")
  else
    return('<div style="page-break-before: always;" />')
}
```

## Matched Cohort

A matched cohort study utilizes specific summary statistics across the group. Recommendations are taken from the following references:

  * Austin PC. A critical appraisal of propensity-score matching in the medical literature between 1996 and 2003. Statist Med 2008; 27: 2037–49. doi:10.1002/sim.3150.
  * Fleiss JL, Levin B, Paik MC. Statistical Methods for Rates and Proportions (3rd edn). Wiley: New York, NY, 2003.
  
The goal is to be able to use tangram defaults as much as possible, while using the following tests.

### For 1:1 matching

| variables | statistical test | Notes|
|:--|:----|:--------|
| Numeric x Cat    | paired Student's t        | `t.test(x=covariate, y=arm, paired=TRUE)`      | 
|                  | Wilcoxon  *(preferred)*     | `wilcox.test(x=covariate, y=arm, paired=TRUE)` |
|                  | Cox proportional hazards models stratifying on matched groups | `survival::coxph(outcome ~ covariate + strata(block), data = m1.final)` |
|                  |                           | useful for time to event analysis |
| Cat X Cat        | McNemar's                 | `mcnemar.test(x=covariate, y=arm)` |
|                  |                           | 2 x 2 cases only. Expects factors |
|                  | Stuart Maxwell chi-squared test | `DescTools::StuartMaxwellTest(x=covariate, y=arm)` |
|                  |                           | for 2 x k polytomous covariates, where k >= 2 |
|                  |                           | expects factors |

### For 1:many matching

| variables | statistical test | Notes|  |
|:--|:----|:--------|
| Numeric x Cat   |   logistic regression with generalized estimated equations | `geepack::geeglm(formula = outcome ~ covariate,` `family = binomial("logit"), data = m2.final, id = block,` `corstr = "independence", zcor = "zcor")` |
|                 | | outcome must be binary numeric (not a factor) |
|                 | | covariate must be numeric |
|                 | | block must be numeric (not a factor) |
| Cat x Cat       | conditional logistic regression | `survival::clogit(outcome ~ covariate + strata(block), data = m2.final)` | Cat x Cat       | Cochran-Mantel-Haenszel chi-squared test | `stats::mantelhaen.test(x=covariate, y=arm, z=block)` |
|                 | | strata with only 1 occurence cause errors, there should be check for this
|                 | | covers 2 x 2 and 2 x >2 polytomous covariates

```{r, include=FALSE}

# Create fake data to work with

set.seed(1)

# patient identifier
study_id <- c(1:100)

# match on these variables
age     <- rnorm(n=length(study_id),mean=10,sd=2)                            # years
sex     <- factor(sample(c(0,1), replace=TRUE, size=length(study_id)))       # M=0, F=1
lang    <- factor(sample(c(0,1,2,3), replace=TRUE, size=length(study_id)))   # 0 = english, 1 = spanish, 2 = french, 3 = other

# outcomes
opioids <- rlnorm(n=length(study_id),0.88,0.67)                              # opioids in mg/kg
los     <- rlnorm(n=length(study_id),3.4,0.67)                               # length of stay in days
nsaid   <- factor(c(rep(0,20),rep(1,80)))                                    # 0 = no nsaid given, 1 = nsaid given
neuro   <- factor(sample(c(0,1,2,3,4), replace=TRUE, size=length(study_id))) # 0 = cervical, 1 = thoracic, 2 = lumbar, 3 = sacral, 4 = other

# comparison / arm
group   <- c(rep(0,70),rep(1,30))                                            # 0 = historical, 1 = prospective cohort

# create data frame
df     <- as_tibble(cbind(study_id,age,sex,lang,opioids,los,nsaid,neuro,group))

# 1:1 example
m1.out    <- matchit(group ~ age + sex + lang, data = df, method = "nearest", ratio = 1)
m1.df     <- match.data(m1.out)
m1.treat  <- as_tibble(df[row.names(m1.out$match.matrix),"study_id"]) %>%
             mutate(block = row_number())
m1.final  <- as_tibble(df[m1.out$match.matrix,"study_id"]) %>%
             mutate(block = row_number()) %>% 
             rbind(m1.treat) %>%
             left_join(m1.df,by="study_id") %>%
             arrange (block,group)

# 2:1 example
m2.out    <- matchit(group ~ age + sex + lang, data = df, method = "nearest", ratio = 2)
m2.df     <- match.data(m2.out)
m2.treat  <- as_tibble(df[row.names(m2.out$match.matrix),"study_id"]) %>%
             mutate(block = row_number())
m2.contr1 <- as_tibble(df[m2.out$match.matrix[,1],"study_id"]) %>%
             mutate(block = row_number())
m2.contr2 <- as_tibble(df[m2.out$match.matrix[,2],"study_id"]) %>%
             mutate(block = row_number())
m2.final  <- m2.treat %>% rbind(m2.contr1,m2.contr2) %>%
             left_join(m2.df,by="study_id") %>%
             arrange (block,group) 

# change factors back to factors
m1.final$group <- factor(m1.final$group, levels=0:1, labels=c("Historical", "Prospective"))
m1.final$sex   <- factor(m1.final$sex, levels=1:2, labels=c("Male", "Female") )
m1.final$lang  <- factor(m1.final$lang, levels=1:4, labels=c("English", "Spanish", "French", "Other"))
m1.final$nsaid <- factor(m1.final$nsaid, levels=1:2, labels=c("No", "Yes"))
m1.final$neuro <- factor(m1.final$neuro, levels=1:5, labels=c("Cervical", "Thoracic", "Lumbar", "Sacral", "Other"))
m2.final$group <- factor(m2.final$group, levels=0:1, labels=c("Historical", "Prospective"))
m2.final$sex   <- factor(m2.final$sex, levels=1:2, labels=c("Male", "Female"))
m2.final$lang  <- factor(m2.final$lang, levels=1:4, labels=c("English", "Spanish", "French", "Other"))
m2.final$nsaid <- factor(m2.final$nsaid, levels=1:2, labels=c("No", "Yes"))
m2.final$neuro <- factor(m2.final$neuro, levels=1:5, labels=c("Cervical", "Thoracic", "Lumbar", "Sacral", "Other"))

attr(m1.final$age,   "label") <- "Age"
attr(m1.final$opioids,"label")<- "Opioids"
attr(m1.final$opioids,"units")<- "mg/kg"
attr(m1.final$los,   "label") <- "Length of Stay"
attr(m1.final$los,   "units") <- "days"
attr(m1.final$group, "label") <- "Group"
attr(m1.final$sex,   "label") <- "Gender"
attr(m1.final$lang,  "label") <- "Language"
attr(m1.final$nsaid, "label") <- "NSAID Given"
attr(m1.final$neuro, "label") <- "Neurological"

attr(m2.final$age,   "label") <- "Age"
attr(m2.final$opioids,"label")<- "Opioids"
attr(m2.final$opioids,"units")<- "mg/kg"
attr(m2.final$los,   "label") <- "Length of Stay"
attr(m2.final$los,   "units") <- "days"
attr(m2.final$group, "label") <- "Group"
attr(m2.final$sex,   "label") <- "Gender"
attr(m2.final$lang,  "label") <- "Language"
attr(m2.final$nsaid, "label") <- "NSAID Given"
attr(m2.final$neuro, "label") <- "Neurological"
```

## Create functions for matching testing

```{r ptest}
psm <- hmisc

psm[['Cell']][['fraction']] <- function(numerator, denominator, format=3, ...) {
  paste0(numerator,' (',render_f(100*numerator/denominator, format),'%)')
}

psm[['Footnote']] = paste("N is the number of non-missing value.",
  "^1^*t*-test.",
  "^2^Wilcoxon signed rank test.",
  "^3^Cox proportional hazards.",
  "^4^Logistic regression with GEE.",
  "^5^Conditional logistic regression.",
  "^6^McNemar's test.",
  "^7^Stuart Maxwell \u03a7^2^ test.",
  "^8^Cochran-Mantel-Haenszel \u03a7^2^ test."
)

mctest.numxcat <- function(rdata, cdata, cell_style, block, pref_test="default", ...)
{
  # get data
  covariate <- rdata
  outcome   <- cdata
  n_matched <- length(block) / length(levels(as.categorical(block)))
  
  # make the df and sort it
  df <- data.frame(
    covariate=as.numeric(covariate),
    outcome=as.numeric(levels(factor(outcome,levels=c(0,1))))[outcome],
                              block=as.numeric(block)
  ) %>% arrange(block,outcome)
  
  p_val <- NA
  ref   <- " "
    
  # first branch point is whether data is matched 1:1 or 1:many
  if (n_matched == 2) {

    # paired Student's t-test
    if (pref_test == "t.test") {
      # run test
      stat <- t.test(x = df$covariate[df$outcome == 0],
                     y = df$covariate[df$outcome == 1],
                     paired = TRUE)
      ref  <- "1"
      
      if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value
      
    # Wilcoxon signed rank test
    } else if (pref_test == "default" || pref_test == "wilcox.test") {
      # run test
      stat <- wilcox.test(x = df$covariate[df$outcome == 0],
                          y = df$covariate[df$outcome == 1],
                          paired=TRUE)
      ref  <- "2"
      
      if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value

    # Cox proportional hazards model stratefied on matched pairs
    } else if (pref_test == "coxph") {
      # run regression
      stat <- survival::coxph(outcome ~ covariate + strata(block), data = df)
      ref  <- "3"
    
      if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value
    }
    
  } else if (n_matched > 2) {
    
    # logistic regression with generalized estimating equations
    if (pref_test == "default" || pref_test == "geeglm") {
      # run regression
      stat <- suppressWarnings(
        geepack::geeglm(formula = outcome ~ covariate,
                        family  = binomial("logit"),
                        data    = df,
                        id      = block,
                        corstr  = "independence",
                        zcor    = "zcor")
      )
      ref <- "4"
      
      if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value[[1]]

    # conditional logistic regression   
    } else if (pref_test == "clogit") {
      # run regression
      stat <- survival::clogit(outcome ~ covariate + strata(block), data = df)
      ref  <- "5"    

      if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value[[1]]
    } 
  }
  
  paste0(cell_style[['p']](p = p_val), "^", ref, "^")
}

mctest.catxcat <- function(rdata, cdata, cell_style, block, ...)
{
  covariate <- as.categorical(rdata)
  outcome   <- as.categorical(cdata)
  block     <- as.categorical(block)
  n_matched <- length(block) / length(levels(as.categorical(block)))
  grid          <- table(covariate, outcome, block, useNA="no")
  validrow      <- which(!apply(grid,1,FUN = function(x){all(x == 0)}))
  validcol      <- which(!apply(grid,2,FUN = function(x){all(x == 0)}))
  validblocks   <- which(!apply(grid,3,FUN = function(x){all(x == 0)}))
  invalidstatum <- which(apply(grid,1,FUN = function(x){sum(ifelse(x==1,1,0))})==1)

  # make the df and sort it
  df <- data.frame(var=as.numeric(covariate), 
                   outcome=as.numeric(levels(factor(outcome,levels=c(0,1))))[outcome],
                   block=as.numeric(block) )
  df <- df %>% arrange(block,outcome)
  
  p_val <- NA
  ref   <- " "
  
  if (n_matched == 2 && length(levels(covariate)) == 2) {
    # McNemar's test

    # x and y must be equal length vectors and have same levels
    # also removed hard-coded outcome levels in cases of other 2-level variables are used
    # (e.g., "M","F" or 1,2 or "No","Yes")
    # code above was somehow stripping a level if the responses were only one level
    # each stats test could have exclusions based on expectations of the underlying
    # function with smarter error handling messages
    stat <- exact2x2::mcnemar.exact(
      x = factor(df$var[df$outcome == levels(outcome)[[1]]], levels = levels(covariate)), 
      y = factor(df$var[df$outcome == levels(outcome)[[2]]], levels = levels(covariate)),
      conf.level=.95)
    ref  <- "6"
    
    # get p value
    if (length(stat) > 1) p_val <- stat$p.value

  } else if (n_matched == 2 && length(levels(covariate)) > 2) {
    
    # Stuart Maxwell chi-squared test
    stat <- StuartMaxwellTest(x=df$var[df$outcome == 0], y=df$var[df$outcome == 1])
    ref  <- "7"

    # get p value
    if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value[[1]]

  } else if (n_matched > 2) {
    
    # Cochran-Mantel-Haenszel chi-squared test
    stat <- if(length(validrow)      < 2 ||
               length(validcol)      < 2 ||
               length(validblocks)   < 1 ||
               length(invalidstatum) > 0) NA else
               mantelhaen.test(covariate,outcome,block)
    ref  <- "8"

    # get p value
    if (length(stat) > 1) p_val <- broom::tidy(stat)$p.value
  }
  
  paste0(cell_style[['p']](p = p_val), "^", ref, "^")
}

mctest <- function(row, col, cell_style, block=NULL, ...)
{
  if(is.null(block)) stop("Block must be specified for matched cohort testing")
  
  if(is.numeric(row$data)  && is.categorical(col$data))
    return(mctest.numxcat(row$data, col$data, cell_style, block, ...))
  if(is.categorical(row$data) && is.categorical(col$data))
    return(mctest.catxcat(row$data, col$data, cell_style, block, ...))
  
  stop(paste("Unsupported comparison for", row$name, "x", col$name, "\nAppears to be", 
    hmisc_data_type(row$data), "X", hmisc_data_type(column$data)))
}
```

`r pagebreak()`

## Results

```{r match11}
# 2-level categories are typically dependent variable in propensity score matching
tangram ( group
            ~ age[1]     # numeric
            + sex        # binary categorical
            + lang       # multi-level categorical
            + opioids[1] # numeric
            + los[1]     # numeric
            + nsaid      # binary categorical
            + neuro,     # multi-level categorical
        data      = m1.final,
        block     = m1.final$block,
        test      = mctest,
        transform = psm,
        digits    = 2,
        style     = "nejm",
        caption   = "Example 1:1 Matching")
```


`r pagebreak()`

```{r table1m}
# 2-level categories are typically dependent variable in propensity score matching
tangram ( group
            ~ age[1]     # numeric
            + sex        # binary categorical
            + lang       # multi-level categorical
            + opioids[1] # numeric
            + los[1]     # numeric
            + nsaid      # binary categorical
            + neuro,     # multi-level categorical
        data      = m2.final,
        block     = m2.final$block,
        test      = mctest,
        transform = psm,
        digits    = 2,
        style     = "nejm",
        caption   = "Match 1:k Example")
```
